---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# ellmer.extensions

<!-- badges: start -->
[![R-CMD-check](https://github.com/xmarquez/ellmer.extensions/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/xmarquez/ellmer.extensions/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`ellmer.extensions` provides backward-compatible entry points for projects that
used its Groq, Gemini, and Anthropic providers before the corresponding features
were added to [ellmer](https://github.com/tidyverse/ellmer):

- `chat_groq_developer()` for Groq structured outputs and batch support.
- `chat_gemini_extended()` for Gemini file-based batches and optional batch
  context caching.
- `chat_anthropic_extended()` for extended or adaptive thinking with structured
  output.
- `openai_costs()` for actual OpenAI organization costs by API key, project, or
  line item.

On current ellmer versions the constructors delegate to native provider
functionality wherever possible. The bundled implementations remain as a
temporary compatibility path for ellmer 0.4.0 and for Gemini batch context
caching. The returned chats work with ellmer generics such as `batch_chat()`,
`batch_chat_structured()`, and `parallel_chat()`.

## Installation

You can install the development version of `ellmer.extensions` from GitHub:

```r
# install.packages("devtools")
devtools::install_github("xmarquez/ellmer.extensions")
```

## Setup

Set the provider keys you use in `.Renviron` (edit with
`usethis::edit_r_environ()`):

```
GROQ_API_KEY=your-groq-key
GEMINI_API_KEY=your-gemini-key
ANTHROPIC_API_KEY=your-anthropic-key
OPENAI_ADMIN_KEY=your-read-only-openai-admin-key
```

(`GOOGLE_API_KEY` also works for Gemini.)

`OPENAI_ADMIN_KEY` is separate from the ordinary key used for model calls. It
should be an Organization Admin API key with read-only access and is needed only
for cost reporting.

## Examples

### Basic Groq chat

```{r}
library(ellmer.extensions)

chat_groq <- chat_groq_developer(
  model = "openai/gpt-oss-20b",
  params = ellmer::params(reasoning_effort = "low")
)
chat_groq$chat("What is the capital of France?")
```

`reasoning_effort` is included in both synchronous and batch request bodies.
Groq's GPT-OSS models otherwise use the provider default (currently `"medium"`).
Arguments supplied through `api_args` are also retained for compatibility and
take precedence over values in `params`.

### Basic Gemini chat

```{r}
library(ellmer.extensions)

chat_gemini <- chat_gemini_extended(model = "gemini-2.5-flash")
chat_gemini$chat("Reply with exactly one word: hello")
```

### Structured output (Groq)

```{r}
type_person <- ellmer::type_object(
  name = ellmer::type_string(),
  age = ellmer::type_integer(),
  city = ellmer::type_string()
)

result <- chat_groq$chat_structured(
  "John is 30 years old and lives in New York City",
  type = type_person
)

str(result)
```

### Batch chat (Groq)

Process multiple requests asynchronously using Groq's batch API:

```{r}
prompts <- list(
  "What is 2 + 2? Reply with only the number.",
  "What is the capital of New Zealand? Reply with only the city."
)

path <- tempfile(fileext = ".json")

chats <- batch_chat(
  chat_groq,
  prompts = prompts,
  path = path,
  wait = TRUE
)

chats
```

### Batch structured output (Groq)

```{r}
type_answer <- ellmer::type_object(
  question = ellmer::type_string(),
  answer = ellmer::type_string()
)

path <- tempfile(fileext = ".json")

data <- batch_chat_structured(
  chat_groq,
  prompts = list("What is 2+2?", "What is the capital of France?"),
  path = path,
  type = type_answer
)

data
```

### Batch chat (Gemini)

Gemini batch processing uses the [file-based Batch API](https://ai.google.dev/gemini-api/docs/batch-api).
Requests are uploaded as a JSONL file, processed asynchronously at 50% of standard
pricing, with a target 24-hour turnaround (many jobs complete faster). Maximum input
file size is 2 GB.

```{r}
prompts <- list(
  "Reply with exactly: ok",
  "Reply with exactly: done"
)

path <- tempfile(fileext = ".json")

# Submit without waiting, then resume later
batch_chat(
  chat_gemini,
  prompts = prompts,
  path = path,
  wait = FALSE
)

batch_chat_completed(chat_gemini, prompts, path)

# Resume and retrieve once completed
chats <- batch_chat(
  chat_gemini,
  prompts = prompts,
  path = path,
  wait = TRUE
)
```

### Parallel chat

Process multiple prompts concurrently (synchronous):

```{r}
parallel_chat(
  chat_groq,
  prompts = list(
    "Tell me one fact about Paris.",
    "Tell me one fact about Wellington.",
    "Tell me one fact about Tokyo."
  )
)
```

## Model discovery

- Groq models: `models_groq()`
- Gemini models: `ellmer::models_google_gemini()`

## OpenAI cost reporting

`openai_costs()` retrieves actual charged amounts from OpenAI's organization
Costs API. It automatically follows pagination and can filter or group reports
by API key ID, project ID, or line item.

```{r}
library(ellmer.extensions)

costs <- openai_costs(
  start_time = Sys.Date() - 30,
  group_by = c("api_key_id", "line_item")
)

sum(costs$amount)
```

API key filters use OpenAI's key identifier (for example, `"key_abc123"`), not
the secret `"sk-..."` value. OpenAI does not currently document an API endpoint
for remaining prepaid credit; view that balance in the organization Billing
page.

## Groq support

Provider model lists and capabilities change frequently. Use `models_groq()` for
the models available to your account and consult Groq's current documentation for
[structured outputs](https://console.groq.com/docs/structured-outputs),
[reasoning](https://console.groq.com/docs/reasoning), and
[batch processing](https://console.groq.com/docs/batch).

## Gemini batch processing

Gemini processing uses the [file-based Batch
API](https://ai.google.dev/gemini-api/docs/batch-api). On current ellmer versions,
ordinary batch submission and result parsing are delegated upstream.
`chat_gemini_extended(cache_ttl = 86400)` retains this package's opt-in context
cache lifecycle for batch system prompts. Use `batch_chat_completed()` to check a
previously submitted job, and `ellmer::models_google_gemini()` for model discovery.

## Anthropic extended thinking with structured output

`chat_anthropic_extended()` preserves the older package interface for combining
thinking with structured output. Current ellmer versions handle this natively;
older versions use this package's `output_config.format` compatibility path.

```{r}
library(ellmer.extensions)

chat <- chat_anthropic_extended(
  model = "claude-sonnet-4-5-20250929",
  params = ellmer::params(reasoning_tokens = 8192)
)

# Structured output works with thinking enabled
type_analysis <- ellmer::type_object(
  score = ellmer::type_number(),
  explanation = ellmer::type_string()
)

result <- chat$chat_structured(
  "Analyze this text for populist rhetoric...",
  type = type_analysis
)

str(result)
```

When `reasoning_tokens` is present in `params`, thinking is automatically
enabled. Without `reasoning_tokens`, `chat_anthropic_extended()` behaves
identically to `ellmer::chat_anthropic()`.

## License

MIT

This package was mostly coded by Claude Code, though everything was reviewed by Xavier Marquez, including all live tests.
